Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-11-18 index.qmd
section:issues; process:write

<!-- 2. But what are limitations of Case V (and the BTL model) -->
However, Case V (and its BTL model counterpart) was originally developed to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp. 269], prioritizing statistical simplicity over precision in trait measurement [@Kelly_et_al_2022, pp. 677]. As a result, its assumptions may not suit applications beyond the psycho-physical contexts for which it was created. Thurstone himself cautioned that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp. 270], acknowledging that some assumptions could prove problematic with more complex, less homogeneous stimuli, such as handwriting or English compositions [@Thurstone_1927a, pp. 374]. Consequently, given that current CJ applications often deal with complex, less homogeneous stimuli, two key assumptions of Case V may not consistently hold in theory or practice: the zero correlation and equal dispersion between stimuli.

@@

<!-- 3. What does the zero correlation between stimuli imply? -->
The assumption of *zero correlation between stimuli* is best illustrated with an example. For instance, when evaluating text quality, this assumption suggests that a judge’s perception of quality in one text does not influence their perception of the same trait in the comparison text. Thurstone attributes this independence to the cancellation of potential judges' biases, driven by opposing and equally weighted factors; commonly known as 'mood' and 'simultaneous contrast' effects, that operate during pairwise comparisons [@Thurstone_1927b, pp. 268]. This cancellation of bias has been mathematically demonstrated by @Andrich_1978, assuming additive biases and a logit scale derived from the BTL model.

<!-- 4. But what are the problems of assuming zero correlation? -->
However, given that evidence has shown judges' quality assessments are influenced by multiple, complex aspects of the texts [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022], it is plausible that the accuracy of these judgments also varies based on factors such as age, culture, education, expertise [@Kelly_et_al_2022, pp. 683], and ultimately, based on the individual judge themselves [@Gill_et_al_2013; @vanDaal_et_al_2017; @vanDaal_2020].

[This means that the cancellation of judges' biases may not hold in theory or practice, and that judges' biases may depend on the stimuli, by means of an interaction (but present) or because the judges works as a confounder ]{style="color:red;"}

@@

<!-- The accuracy of judgment may vary by subject and by judge [@Gill_et_al_2013; @vanDaal_et_al_2017; @vanDaal_2020] -->

<!-- What we are saying is that the judges' effects depend on the stimuli, by means of an interaction (but present) or because the judges works as a confounder. This, in the best scenario, cause less precise estimates, and in the worst scenario, biased estimates. possibility of bias is hinted in @Bramley_2008, and later mentioned again in @Kelly_et_al_2022, and some evidence is found in @Pollitt_et_al_2003, @vanDaal_et_al_2017 and @Gijsen_et_al_2021 -->

<!-- Such differences may not be detected through analyses of bias and misfit. McMahon and Jones (2015) found differences between students and teachers in what was valued in assessments of understanding of a chemistry experiment. Where students prioritised factual recall, teachers prioritised scientific explanation. Both groups produced highly reliable scales (reliability estimates of .893 for students, and .874 for teachers). Although, in this case, a clear argument as to whose consensus should take priority can be made, it demonstrates the existence of different, but equally reliable, sets of consensuses depending on who is asked. [@Kelly_et_al_2022, pp. 683] -->

<!-- this also translates into the idea that stimuli are the main focus of estimation and analysis, but what happens when the focus of analysis is the individuals that generated those stimuli. Meaning there is an amount of correlation that it is not accounted for. Use example of Boonen!! -->

@@

>>>>

<!-- 2. But what are limitations of Case V (and the BTL model) -->
However, Case V was originally developed to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp. 269], prioritizing statistical simplicity over precision in trait measurement [@Kelly_et_al_2022, pp. 677]. As a result, its assumptions may not suit applications beyond the psycho-physical contexts for which it was created. Thurstone himself cautioned that its use "should not be made without experimental test" [@Thurstone_1927b, pp. 270], acknowledging that some assumptions could prove problematic with complex traits or less homogeneous stimuli [@Thurstone_1927a, pp. 374]. Consequently, given that current CJ applications often deal these types of traits and stimuli, two key assumptions of Case V may not consistently hold in theory or practice: the zero correlation and equal dispersion between stimuli.

@@

<!-- 3. What does the zero correlation between stimuli imply? -->
The assumption of *zero correlation between stimuli* is best illustrated with an example. For instance, when evaluating text quality, the assumption suggests that a judge’s perception of quality in one text does not influence the perception of the same trait in the comparison text. Thurstone attributes this independence to the cancellation of potential judges' biases, driven by opposing and equally weighted factors that operate during pairwise comparisons, called 'mood' and 'simultaneous contrast' effects [@Thurstone_1927b, pp. 268]. This cancellation of bias has been mathematically demonstrated by @Andrich_1978, assuming discriminal processes with additive biases and a logit scale derived from the BTL model.

<!-- 4. But what are the problems of assuming zero correlation? -->
However, two types of scenarios make it plausible that potential judges' biases may not cancel: those involving complex traits and less homogeneous stimuli, and those where judges differ in what they value in their assessments. In the first scenario, Thurstone noted that for complex traits and non-homogeneous stimuli, such as handwriting or English compositions, CJ data might not align with the assumptions of Case V [@Thurstone_1927a, pp. 374]. This insight likely extends to other similarly complex traits and stimuli. In the second scenario, evidence indicates that judges' assessments are influenced by multiple, intricate aspects of the stimuli [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. Moreover, factors like age, culture, education, expertise [@Kelly_et_al_2022, pp. 683], and even individual differences among judges [@Gill_et_al_2013; @vanDaal_et_al_2017; @vanDaal_2020] can influence judgment accuracy. These scenarios can ultimately result in non-additive biases that resist cancellation, driven by the characteristics or location of stimuli within the trait continuum.

<!-- seen as a strength of the method [@vanDaal_et_al_2016], and commonly referred to as "the wisdom of crowds" [@Surowiecki_2004] -->

<!-- This means that the cancellation of judges' biases may not hold in theory or practice, and that judges' biases may depend on the stimuli, by means of an interaction (but present) or because the judges works as a confounder -->

@@

<!-- 5. But there is another problem associated with the zero correlation that no one is talking -->
[This also translates into the idea that stimuli are the main focus of estimation and analysis, but what happens when the focus of analysis is the individuals that generated those stimuli. Meaning there is an amount of correlation that it is not accounted for. Use example of Boonen!! ]{style="color:red;"}

@@

<!-- Such differences may not be detected through analyses of bias and misfit. McMahon and Jones (2015) found differences between students and teachers in what was valued in assessments of understanding of a chemistry experiment. Where students prioritised factual recall, teachers prioritised scientific explanation. Both groups produced highly reliable scales (reliability estimates of .893 for students, and .874 for teachers). Although, in this case, a clear argument as to whose consensus should take priority can be made, it demonstrates the existence of different, but equally reliable, sets of consensuses depending on who is asked. [@Kelly_et_al_2022, pp. 683] -->
<!-- What we are saying is that the judges' effects depend on the stimuli, by means of an interaction (but present) or because the judges works as a confounder. This, in the best scenario, cause less precise estimates, and in the worst scenario, biased estimates. possibility of bias is hinted in @Bramley_2008, and later mentioned again in @Kelly_et_al_2022, and some evidence is found in @Pollitt_et_al_2003, @vanDaal_et_al_2017 and @Gijsen_et_al_2021 -->
